---
title: R Cheatsheet for Gov 51
date: 2020-09-11
summary: "Guide to the R techniques used in Gov 51"
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set()
options(width = 90)
```

# Week 0

## Vectors 

### Create a vector 

Use the `c()` function for concatenate. 

```{r}
x <- c(2.2, 6.2, 1.2, 5.5, 20.1)
```

### Creating a vector from a sequence

The `seq()` function will create a sequence from a starting point to an ending point. If you specify the `by` argument, you can skip values. For instance, if we wanted a vector of all presidential election years from 1788 until 2020, we would use:

```{r}
pres_elections <- seq(from = 1788, to = 2020, by = 4)
pres_elections
```

If we want to a sequence from `a` to `b` counting by 1s, we can use the `a:b` shorthand:

```{r}
2014:2020
```

### Length of a vector

```{r}
length(x)
```

### Summarizing a numeric vector

The `summary` function will give you some summary statistics about a vector:

```{r}
summary(pres_elections)
```

### Means, median, sums, min, max

There are various simple summary functions for specific summaries. 

```{r}
mean(x)
median(x)
sum(x)
```

### Ranges of a vector

You can easily get the smallest and largest values of a vector with `min`, `max`, and `range`. 

```{r}
min(x)
max(x)

## get both in one vector
range(x)
```

### Subsetting a vector

To access a single element of a vector by position in the vector, use the square brackets `[]`:

```{r}
x[2]
```

If you want to access more than one element of a vector, put a vector of the positions you want to access in the brackets:

```{r}
x[c(2, 5)]
```

### Subsetting by a logical vector

The last two approaches selected subsets by position ("give me the 2nd and 5th element of this vector"). Many times, we will want to select the values of a vector that meet some logical condition. To achieve this, we can put a logical vector **the same length as the vector being subsetted** in the square `[]` brackets with `TRUE` or `FALSE` indicating if each entry should or should not be included in the subset. For instance, we can recreate the `x[c(2, 5)]` subset with the following:

```{r}
x[c(FALSE, TRUE, FALSE, FALSE, TRUE)]
```

More usefully, we can create logical vectors from expressions to get all values from the vector that meet some condition:

```{r}
x[x > 5]
``` 

**Gotcha**: if your logical subsetting vector that is shorter than the length of the vector being subsetted, R will implicitly repeat the logical vector to the right length. So these two commands give the same output:

```{r}
x[c(TRUE, FALSE)]
x[c(TRUE, FALSE, TRUE, FALSE, TRUE)]
```

If your logical subsetting vector is longer than the vector being subsetted, you will get `NA` values for the values beyond the correct length:

```{r}
x[c(TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE)]
```

### Subsetting accidents

If you try to access an element past the length of the vector, it will return a missing value `NA`:

```{r}
x[10]
```


If you accidentally subset a vector by `NA` (the missing value), you get the vector back with all its entries replaced by `NA`:

```{r}
x[NA]
```

Sometimes you'll notice that a vector or data frame you've been messing with all of a sudden has lots of missing values (maybe whole rows of them). Usually the culprit in that case is some errant missing value in a subset. 

### Subsetting to the first 5 values

```{r}
pres_elections[1:5]
head(pres_elections, n = 5)
```


### Subsetting to the last value in a vector

The `length` function gives us the number of positions in the vector and so it is also the last position. You can also use the tail function. 

```{r}
pres_elections[length(pres_elections)]
tail(pres_elections, n = 1)
```


### Modifying values in a vector 

Let's say you want to modify one value in your vector. You can combine the square bracket subset `[]` with the assignment operator `<-` to replace a particular value:

```{r}
x
x[3] <- 50.3
x
```

You can replace multiple values at the same time by using a vector for subsetting:

```{r}
x
x[1:2] <- c(-1.3, 42)
x
```

If the replacement vector (the right-hand side) is shorter than what you are assigning to (the left-hand side), the values will "recycle" or repeat as necessary:

```{r}
x[1:2] <- 3.2
x

x[1:4] <- c(1.2, 2.4)
x
```

**Gotcha:** if the right-hand side length is not a multiple of the left-hand side length, R will do the replacement and truncate things correctly, but will complain and give a warning:

```{r}
x[1:3] <- c(0.5, 1.5)
x
```

Usually if you get this warning in Gov 51, something has gone wrong and you should check your code. (That's a more general rule about warnings in Gov 51.)

### Arithmetic on vectors

You can perform simple arithmetic on vectors and it will apply to each element:

```{r}
x
x * 10
```

You can also take apply arithmetic on multiple vectors that are the same lengths. Suppose we have two vectors of test scores for 5 students. We can calculate the change in their scores with:

```{r}
score_1 <- c(88, 95, 92, 75, 82)
score_2 <- c(90, 92, 97, 81, 85)
score_2 - score_1
```

Of course, we could also save that vector for future use:

```{r}
score_change <- score_2 - score_1
mean(score_change)
```

### Character vectors

You can also create a vector of characters (words, letters, punctuation, etc):

```{r}
jedi <- c("Yoda", "Obi-Wan", "Luke", "Leia", "Rey")
```

Note for vectors, you cannot mix characters and numbers in the same vector. If you add a single character element, the whole vector gets converted. 

```{r}
## output is numeric
x

## output is now character
c(x, "hey")
```

## Data frames

### Reading in CSV files

You can import data into R using the `read.csv()` function. 

```{r, eval = FALSE}
my_data <- read.csv("data/mydata.csv")
```

**Troubleshooting**: if you get an error trying to load data, double check the name of the file you are trying to load and make sure that the csv file exists where you think it does. You can always look at the Files pane of RStudio to look in a data folder to see what the contents are and what the names of the files are you are trying to import. 

### Reading in RData files

You can read files with an `.RData` extension with the `load()` function:

```{r, eval = FALSE}
load("data/mydata.RData")
```

With the `load()` function, you don't actually assign its output to anything. An `.RData` file contains R objects (vectors, data frame, etc) and when you load it, those objects are dumped into your R environment. 

### Dimensions of a data frame

The `cars` data frame is built into R and so you can access it without loading any files. To get the dimensions, you can use `dim()`, `nrow()`, and `ncol()`.

```{r}
dim(mtcars)
nrow(mtcars)
ncol(mtcars)
```

### Accessing variables/columns

You can grab each column/variable from the data frame use the `$`, turning it into a vector:

```{r}
mtcars$wt
```

You can now treat this just like a vector, with the subsets and all. 

```{r}
mtcars$wt[1]
```

### Subsetting a data frame


You can subset data frame using the square brackets `[]`, but things are somewhat more complicated. Here, you have to use the comma: `[,]`. Before the comma comes the subsets for rows and after the comma subsets the columns. 

```{r}
mtcars[, "wt"]
```

Grab the first 5 rows:

```{r}
mtcars[1:5,]
```

Grab the first 3 columns:

```{r}
mtcars[,1:3]
```

### Summarizing a data frame

If you call `summary()` a data frame, it produces applies the vector version of the summary command to each column:

```{r}
summary(mtcars)
```

### Subset to the first/last k rows of a data frame

```{r}
head(mtcars)
tail(mtcars)
```

# Week 1

For this section, we'll use the `social` data from QSS, which is the social pressure mailer for voter turnout experiment. 

```{r}
data("social", package = "qss")
```

## Getting counts and cross-tabs

### Get counts of each value in a vector

When passed a single vector, the `table` function will return the number of entries that take each unique value. That is, it gives you the counts of each unique value. 

```{r}
table(social$primary2004)
```

### Cross-tabulation

When passed two vectors, the `table` function will return counts for each combination of the two vectors:

```{r}
table(social$primary2004, social$primary2006)
```

You can give these entries names to produce slightly nicer tables that have labels:

```{r}
table(Turnout2004 = social$primary2004, Turnout2006 = social$primary2006)
```

If you want to add a space in the label, you should put it in quotes:

```{r}
table("Turnout in 2004 Primary" = social$primary2004,
      "Turnout in 2006 Primary" = social$primary2006)
```

Adding a third vector will give you a cross-tab for each unique value of the third variable.

```{r}
table(social$primary2004, social$primary2006, social$sex)
```

### Converting counts to proportions using `prop.table`

You can convert any `table()` output from counts to proportions by wrapping the call in `prop.table`:

```{r}
prop.table(table(social$primary2006))
```

## Controlling output

### Rounding numbers 

R often gives you way more significant digits than you need to display. You can always pass a vector or number to `round()` to display/print only a certain number of digits:

```{r}
round(prop.table(table(social$primary2006)), digits = 2)
```

This is useful in R Markdown files when you want to access a particular value in text as `` `r "\x60r mean(social$primary2006)\x60"` `` which would be `r mean(social$primary2006)`. You can instead use `` `r "\x60r round(mean(social$primary2006), 2)\x60"`  `` to get `r round(mean(social$primary2006), 2)`. 

## Logicals


### Logical vectors

Logical vectors are just vectors that only contain the special R values `TRUE` or `FALSE`. 

```{r}
c(TRUE, FALSE, TRUE, TRUE, FALSE)
```





### TRUE/FALSE are case sensitive 

You could but never should shorten `TRUE` to `T` and `FALSE` to `F`. It's easy for this shortening to go wrong so better just to spell out the full word. Also not that this is case-sensitive, and this will produce an error:

```{r, error = TRUE}
true
True
false
```

### Converting logicals to numerics

Like numeric and character vectors, you cannot mix these types with other values. But if you treat a logical vector like a numeric and apply a function to it, R will usually silent convert the logical to a numeric with `TRUE` being assigned to `1` and `FALSE` being assigned to `0`. You can see this explicitly by passing a logical vector to the `as.numeric` function:

```{r}
as.numeric(c(TRUE, FALSE, TRUE, TRUE, FALSE))
```

### The number and proportion of TRUE values

Because of the conversion mentioned above, passing a logical vector to the `sum()` function (which adds all of the entries together) will return the number of true values in the vector:

```{r}
sum(c(TRUE, FALSE, TRUE, TRUE, FALSE))
```

The `mean()` function will give the proportion of `TRUE` values in the vector:

```{r}
mean(c(TRUE, FALSE, TRUE, TRUE, FALSE))
```

### and/or statements with logicals


We often combine logical statements using AND (`&`) and OR (`|`) in R. For AND statements, both expressions have to be true for the whole expression to be true:

- `TRUE & FALSE`, `FALSE & TRUE`, and `FALSE & FALSE` are `FALSE`
- `TRUE & TRUE` is `TRUE`

For OR statements, either statement being true makes the whole expression true:

- `TRUE | FALSE`, `FALSE | TRUE`, and `TRUE | TRUE` are `TRUE`
- `FALSE | FALSE` is `FALSE`

You can also use these to compare two logical vectors:

```{r}
## Apply "or" statement to each pair of elements
c(TRUE, FALSE, TRUE, TRUE, FALSE) | c(TRUE, TRUE, FALSE, FALSE, FALSE)

# Apply "and" statement to each pair of elements
c(TRUE, FALSE, TRUE, TRUE, FALSE) & c(TRUE, TRUE, FALSE, FALSE, FALSE)
```

### Comparing objects


There are several *relational operators* that allow us to compare objects in R. The most useful of these are the following:

- `>` greater than, `>=` greater than or equal to
- `<` less than, `<=` less than or equal to
- `==` equal to
- `!=` not equal to

When we use these to compare two objects in R, we end us with a logical object. You can also compare a vector to a particular number.

```{r}
pres_elections > 2000
```


### Subset using logical vectors

You can use logical vectors to subset vectors to meet certain conditions:

```{r}
pres_elections[pres_elections > 2000]
```

### Subsetting data frame using `subset`

The `subset` function creates subsets of a data frame that meet certain conditions:

```{r}
control <- subset(social, subset = messages == "Control")
head(control)
```

Note that you don't have to use the `resume$firstname` syntax in the subset argument here. The function smartly knows to find `firstname` in the `resume` data frame. 

### `subset` with multiple conditions

```{r}
control_males <- subset(social, subset = messages == "Control" & sex == "male")
head(control_males)
```

# Week 2

### Two possible values using `ifelse`

`ifelse` is a function that will take a logical vector and create a new vector that takes two different values when the entry in the logical vector is either `TRUE` or `FALSE`. The first argument is the logical vector, the second is what the entries of the new vector should be if the logical entry is `TRUE` and the third is what the entries of the new vector should be when the logical entries are `FALSE. `

```{r}
ifelse(c(TRUE, FALSE, TRUE, TRUE, FALSE), 100, -100)
```

You can also create character vectors this way:

```{r}
ifelse(c(TRUE, FALSE, TRUE, TRUE, FALSE), "yessss", "this ain't it")
```

### Creating vectors based on several conditions

If you want to create a vector that has more than two levels based on a several logical conditions, you can do this with subsetting. For example, let's bin `social$ages` into four groups based on their rough generation labels.

- `"Greatest"` for those born in 1927 and before
- `"Silent"` for those born between 1928 until 1945
- `"Boomers"` for those born between 1946 until 1964
- `"GenX"` for those born between 1965 and 1980
- `"Millenial"` those born in 1981 and after

```{r}
## creates a column of NA placeholder values with the name "generation"
social$gen <- NA

## replace each subset separately
social$gen[social$yearofbirth <= 1927] <- "Greatest"
social$gen[social$yearofbirth >= 1928 & social$yearofbirth <= 1945] <- "Silent"
social$gen[social$yearofbirth >= 1946 & social$yearofbirth <= 1964] <- "Boomers"
social$gen[social$yearofbirth >= 1965 & social$yearofbirth <= 1980] <- "GenX"
social$gen[social$yearofbirth >= 1981] <- "Millenial"

table(social$gen)
```


### Calculating the mean within groups using `tapply`

An efficient way to do this task would be to use the `tapply(X, INDEX, FUN)` function, which allows you to compute a function (`FUN`) on subsets of the data (`X`) defined by a factor variable (`INDEX`). For instance, to calculate the mean of the `primary2006` variable within levels the treatment `messages` variable we use:

```{r}
tapply(social$primary2006, social$messages, mean)
```

# Week 3

## Visualization

### Barplot

First we create a basic barplot with defaults:

```{r barplot-plain}
barplot(table(social$primary2006))
```

Now we can also add in a bunch of arguments to make it more attractive:

```{r barplot}
barplot(
  table(social$primary2006),
  names.arg = c("Didn't Vote", "Voted"), ## group labels, L to R
  main = "Distribution of Turnout in 2006 Primary", ## main title
  ylab = "Number of Respondents", ## y-axis label
  col = "lightblue",  ## color of the bars
  border = NA ## remove black border
)
```


### Histograms

Again, a plain histogram with just the defaults except `freq = FALSE` to ensure the histogram is giving us density.

```{r hist-plain}
social$age <- 2006 - social$yearofbirth
hist(social$age, freq = FALSE)
```

We can also make it fancy using various arguments. You may only want to use some of these. 

```{r hist-fancy}
hist(
  social$age,
  freq = FALSE,
  main = "Distribution of Subject Ages",
  xlab = "Age",
  col = "lightblue",
  border = "white",
  las = 1,  ## make the y-axis tick labels horizontal
  cex.axis = 0.8 ## size of axis tick labels relative to 1
)
```
The `cex.axis` argument below is expressed as a fraction of the default size. So, setting this to 1 would be the default text size, and `0.8` makes it 20% smaller. This is helpful when we use `las=1` to make the axis labels horizontal.

### Boxplots

Here is the basic boxplot code with the defaults:

```{r boxplot-plain}
boxplot(social$age)
```

We can also add in many of the arguments from the other plotting commands:

```{r boxplot-fancy}
boxplot(
  social$age,
  main = "Distribution of Subject Age",
  ylab = "Age",
  col = "lightblue",
  las = 1,
  border = "grey20"
)
```

## Missing data

### Calculating means of vectors with missing values

To calculate the mean or sum of a vector with missing elements, using `na.rm = TRUE` as the final argument. 

```{r}
x <- c(1:5, NA)
x

mean(x)

mean(x, na.rm = TRUE)

sum(x)

sum(x, na.rm = TRUE)
```

### Testing if vector entries are missing or not

You can test if the entries of a vector are missing using the `is.na()` function:

```{r isna}
x
is.na(x)
```

You can use the NOT operator `!` to give you a vector that is `TRUE` when the vector is not missing:

```{r notna}
!is.na(x)
```


### Removing `NA` values from a vector or data frame

There are a couple of ways to subset to a vector without the missing values. 

```{r remove-na}
x

na.omit(x)

x[!is.na(x)]
```

When applied to a data frame `na.omit()` removes any rows of the data frame that contain any missing values. 

# Week 4

## Summarizing bivariate relationships

For this week, we'll load the `STAR` data from the QSS package:

```{r}
data(STAR, package = "qss")
```

### Scatterplots

To create a basic scatterplot, simply provide the variables corresponding the x and y axes:

```{r}
plot(x = STAR$g4math, y = STAR$g4reading)
```

Of course, we can make these look quite a bit nicer:

```{r}
plot(x = STAR$g4math, y = STAR$g4reading,
     xlab = "Grade 4 Math Score", ylab = "Grade 4 Reading Score",
     las = 1, main = "Reading and Math Score Relationship",
     col = "steelblue2")
```

You can change the shape of the points in the scatterplot using the `pch` (plotting character) argument. For instance, `pch = 19` uses filled dots:

```{r}
plot(x = STAR$g4math, y = STAR$g4reading,
     xlab = "Grade 4 Math Score", ylab = "Grade 4 Reading Score",
     las = 1, main = "Reading and Math Score Relationship",
     col = "steelblue2",
     pch = 19 ## filled points
     )
```

### Making points more transparent in scatterplots

One thing to note is that when points are plotted over one another, solid points can look like a blob. In those cases, it might be reasonable to make the points more transparent. You can achieve this by using the `adjustcolor()` function in your `col` argument like so:

```{r}
plot(x = STAR$g4math, y = STAR$g4reading,
     xlab = "Grade 4 Math Score", ylab = "Grade 4 Reading Score",
     las = 1, main = "Reading and Math Score Relationship",
     col = adjustcolor("steelblue2", alpha = 0.5),
     pch = 19
     )
```

Here, setting the value of `alpha` in `adjustcolor` will change how transparent (0 being non-visible and 1 being full opaque). But now we can tell when points are plotted over one another more clearly. 


### Correlation

To take the correlation between two vectors of the same length, use the `cor()` function. 

```{r}
cor(STAR$g4math, STAR$g4reading)
```

Oops, there are missing values. Unlike other functions, you can't use `na.rm = TRUE` here and instead you want to use `use="pairwise"`:

```{r}
cor(STAR$g4math, STAR$g4reading, use = "pairwise")
```

### Testing if entries of a vectors are equal to one of several values

In the `STAR` data, there is a variable `yearssmall` that describes how many years a student was in a small class type. This can take on 5 different values:

```{r}
head(STAR$yearssmall)
table(STAR$yearsmall)
```

Maybe we want to compare units who are only have 0 or 1 year of small classes. We could do this with an OR statement:

```{r}
STAR$zero_or_one <- ifelse(STAR$yearssmall == 0 | STAR$yearssmall == 1, 1, 0)
head(STAR$zero_or_one)
```

This is a bit clunky, especially if we want to test for multiple categories. It is slightly easier to use the `a %in% b` syntax, which will check every entry of `a` to see if it is in `b` and give `TRUE`, otherwise it will give `FALSE`:

```{r}
STAR$zero_or_one <- ifelse(STAR$yearssmall %in% c(0, 1), 1, 0)
head(STAR$zero_or_one)
```

### Boxplots by group

You can create boxplots by group using a formula: `y ~ x` where `y` is the name of the variable you want to plot in the boxplot and `x` is the name of the grouping variable. You also need to pass the name of the data set to `data`.

```{r}
boxplot(g4math ~ classtype, data = STAR)
```

Of course, we can make this much nicer in appearance using the usual tricks. Note that the class size variable measures type of kindergarten class and takes on these values: small = 1, regular = 2, regular with aid = 3. 

```{r}
boxplot(g4math ~ classtype, data = STAR,
        names = c("Small", "Regular", "Regular w/ Aid"),
        xlab = "Classroom Type",
        ylab = "Grade 4 Math Scores",
        main = "Math Scores by Classroom Type",
        las = 1, col = "steelblue2")
```


### Removing the box in a base R plot

For almost any base R plot command, you can remove the box that the plot sits in using the `frame = FALSE` argument:

```{r}
boxplot(g4math ~ classtype, data = STAR,
        names = c("Small", "Regular", "Regular w/ Aid"),
        xlab = "Classroom Type",
        ylab = "Grade 4 Math Scores",
        main = "Math Scores by Classroom Type",
        las = 1, col = "steelblue2",
        frame = FALSE)
```


### Q-Q plots

Use the `qqplot()` function to produce Q-Q plots:

```{r}
qqplot(x = STAR$g4math, y = STAR$g4reading,
       xlab = "Grade 4 Math Score",
       ylab = "Grade 4 Reading Score",
       las = 1,
       col = "steelblue2")
```


### Producing lines on plots

You can produce straight lines on plots using the `abline()` command. There are three different ways to use it:

- `abline(v = 0)`: this will draw a vertical line at x = 0. 
- `abline(h = 0)`: this will draw a horizontal line at y = 0. 
- `abline(a = 0, b = 1)`: this will draw a straight line with y-intercept 0 and slope 1 (`y = x`). 

We can use this in the Q-Q plot to draw a reference line:

```{r}
qqplot(x = STAR$g4math, y = STAR$g4reading,
       xlab = "Grade 4 Math Score",
       ylab = "Grade 4 Reading Score",
       las = 1,
       col = "steelblue2")
abline(a = 0, b = 1)
```

You can also change the type of line produced using the `lty` argument ("line type"). This [blog post](https://www.benjaminbell.co.uk/2018/02/quick-guide-to-line-types-lty-in-r.html) has a good overview of the different options. You can also pass a `col` argument to change the color of the line and `lwd` to change the thickness (where 1 is the default and 2 is double width, etc)

```{r}
qqplot(x = STAR$g4math, y = STAR$g4reading,
       xlab = "Grade 4 Math Score",
       ylab = "Grade 4 Reading Score",
       las = 1,
       col = "steelblue2")
abline(a = 0, b = 1, lty = 2, col = "grey50", lwd = 3)
```

# Week 5

## Loops and conditionals

### For loops

When coding for data analysis, we often have to perform repetitive tasks that change only in small ways. You could simply copy and paste the code for one of these task and manually change the code as needed, but this will get tedious very quickly. It will also make the code easier to read. 

For example, suppose we wanted to calculate the mean of each column of the `STAR` data. We could achieve this by simply creating the vector by hand:

```{r}
star_means <- c(mean(STAR$race, na.rm = TRUE),
                mean(STAR$classtype, na.rm = TRUE),
                mean(STAR$yearssmall, na.rm = TRUE),
                mean(STAR$hsgrad, na.rm = TRUE),
                mean(STAR$g4math, na.rm = TRUE),
                mean(STAR$g4reading, na.rm = TRUE))
star_means
```

This is a lot of code that does the same thing over and over again. Instead, we can use a loop to perform the same calculate *applied* to each column. To do this, we first need to create the placeholder vector where we will put all of the means, for which we use the `rep` function that creates a vector of a certain (or values) repeated some number of times. Then we start the loop using the `for (i in X)` construction. The `X` is a vector of values to iterate over and the `i` is the name of the iterator, a local variable in the loop that will give back current value of `X` in this iteration of the loop. Usually, `X` is of the form `1:n`. 

```{r}
star_means <- rep(NA, times = ncol(STAR))
for (i in 1:ncol(STAR)) {
  star_means[i] <- mean(STAR[, i], na.rm = TRUE)
}

star_means
```

We iterate over the columns of `STAR` so we use `1:ncol(STAR)` in the setup of the loop and iterate through the column numbers/positions. We then subset the STAR data to that column and take its mean, assigning it to the `i`th position in the holder. 

### Printing output to the R console

You can print something to the R console using the `cat()` command. Note that you need to add at least one `\n` at the end of the string so that the next thing to print will be on the next line. 

```{r}
cat("hello world!\n")
```


### if statements

If statements are a powerful computing tool that allow us to execute code only if some conditions hold.  We might implement it like this:

```{r, eval = FALSE}
if (condition) {
  [code to evaluate]
}
```

For instance, let's have R tell us if we have a negative number:

```{r if, echo = -1}
x <- -4
if (x < 0) {
  cat("Hey! That's a negative!\n")
}
cat("Wow, we've reached the end!\n")
```

Compare this to the same code, but given a positive number:

```{r}
x <- 4
if (x < 0) {
  cat("Hey! That's a negative!\n")
}
cat("Wow, we've reached the end!\n")
```


### if/else statements

You can use the `if {} else {}` construct to have a set of fallback code that executes if the condition is not met. Continuing our theme:

```{r}
x <- -4
if (x < 0) {
  cat("Hey! That's a negative!\n")
} else {
  cat("Whoa! It's positive!\n")
}
```


```{r}
x <- 4
if (x < 0) {
  cat("Hey! That's a negative!\n")
} else {
  cat("Whoa! It's positive!\n")
}
```


## Linear regression

We'll use the leader assassinations data from the QSS book and the lectures.

```{r}
data(leaders, package = "qss")
```


### Fitting a linear regression model

First, we'll regress the polity (aka democracy) score after the assassination attempt on the same score before:

```{r}
fit <- lm(polityafter ~ politybefore, data = leaders)
fit
```


### Obtaining the vector of estimated regression coefficients

```{r}
coef(fit)
```


### Obtaining a vector of fitted/predicted values from the regression

```{r}
head(fitted(fit))
```


### Adding a regression line to a plot

You can use the `abline()` function with the `lm()` output to add a regression line to a plot:

```{r}
plot(x = leaders$politybefore, y = leaders$polityafter,
     pch = 19, col = adjustcolor("steelblue2", alpha = 0.5),
     frame = FALSE, xlab = "Polity Before", ylab = "Polity After")
abline(fit, lwd = 3, col = "indianred")
```


### Obtaining the vector of residuals

```{r}
head(residuals(fit))
```
